---
title: "Healthcare Provider Fraud Detection"
author: "Cotiviti Intern Assessment"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this document we will analyze the healthcare provider data. Main goal of this analysis is to create a logistic regression model
to predict whether a provide is fradulent or not. In this analysis we will first identify the multicollinearity betweeen the features 
we engineered using VIF. Later we will use ridge-logistic regression to create our model.

```{r}
X_train <- read.csv('X_train.csv')
X_val <- read.csv('X_val.csv')
X_test <- read.csv('X_test.csv')

y_train <- read.csv('y_train.csv')
y_val <- read.csv('y_val.csv')
y_test <- read.csv('y_test.csv')
```

## Imputation

NA values in features like `avg_length_of_stay` and `num_unique_procedure_codes` mean the provider had no inpatient claims or billed no procedures — so 0 is the correct fill, not the column mean.

```{r}
X_train[is.na(X_train)] <- 0
X_val[is.na(X_val)]     <- 0
X_test[is.na(X_test)]   <- 0
```

## Correlation Heatmap

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}
library(ggplot2)

X_num <- X_train[, sapply(X_train, is.numeric)]
C <- cor(as.matrix(X_num))

C_df      <- as.data.frame(as.table(C))
names(C_df) <- c("Var1", "Var2", "value")
C_df$high <- abs(C_df$value) > 0.95 &
             as.character(C_df$Var1) != as.character(C_df$Var2)

ggplot(C_df, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_point(data = C_df[C_df$high, ],
             shape = 4, size = 2, color = "black") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "tomato",
                       midpoint = 0, limits = c(-1, 1), name = "cor") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
        axis.text.y = element_text(size = 6)) +
  labs(title = "Feature Correlation Matrix  (X = |cor| > 0.95)",
       x = NULL, y = NULL)

idx <- which(abs(C) > 0.95 & upper.tri(C), arr.ind = TRUE)
if (nrow(idx) > 0) {
  high_df <- data.frame(
    Feature1    = rownames(C)[idx[, 1]],
    Feature2    = colnames(C)[idx[, 2]],
    Correlation = round(C[idx], 3)
  )
  knitr::kable(high_df, caption = "Feature pairs with |cor| > 0.95")
}
```

## Multicollinearity Check (VIF Before Ridge)

VIF values above 5 indicate problematic multicollinearity. We check this before
applying ridge regression to motivate the use of regularization.

```{r, message=FALSE, warning=FALSE}
# compute VIF from correlation matrix — equivalent to regression-based VIF
# but instant (no model fitting required)
X_num <- X_train[, sapply(X_train, is.numeric)]
C <- cor(as.matrix(X_num))
vif_before <- diag(solve(C))
names(vif_before) <- colnames(X_num)

par(mar = c(9, 4, 4, 2))
vif_sorted <- sort(vif_before, decreasing = TRUE)
bp <- barplot(vif_sorted, names.arg = rep("", length(vif_sorted)),
              main = "VIF Before Ridge Regression",
              ylab = "VIF", col = ifelse(vif_sorted > 10, "tomato", "steelblue"))
text(bp, par("usr")[3] - 0.3, labels = names(vif_sorted),
     srt = 45, adj = 1, xpd = TRUE, cex = 0.55)
abline(h = 10, col = "darkred", lty = 2)
```

## Ridge Logistic Regression

```{r, message=FALSE, warning=FALSE}
library(glmnet)
library(pROC)

# drop highly correlated columns before fitting
drop_cols <- c("total_deductible_paid", "avg_deductible_paid",
               "num_unique_attending_physicians", "num_unique_procedure_codes")

keep_num <- function(df) {
  df[, sapply(df, is.numeric) & !names(df) %in% drop_cols]
}

X_tr  <- as.matrix(keep_num(X_train))
X_vl  <- as.matrix(keep_num(X_val))
X_te  <- as.matrix(keep_num(X_test))

train_means <- colMeans(X_tr)        # mean of each column (feature)
train_sds   <- apply(X_tr, 2, sd)    # sd of each column (feature)

X_tr_scaled <- scale(X_tr, center = train_means, scale = train_sds)
X_vl_scaled <- scale(X_vl, center = train_means, scale = train_sds)
X_te_scaled <- scale(X_te, center = train_means, scale = train_sds)

fit <- glmnet(X_tr_scaled, as.numeric(y_train[[1]]),
              family = "binomial", alpha = 0, lambda = NULL, standardize = FALSE)

# predict on validation set for every lambda
val_preds <- predict(fit, newx = X_vl_scaled, type = "response")
# compute AUC for each lambda and pick the best
aucs <- apply(val_preds, 2, function(p) auc(roc(as.numeric(y_val[[1]]), p, quiet = TRUE)))
best_lambda <- fit$lambda[which.max(aucs)]

cat("Best lambda:", best_lambda, "\nBest val AUC:", max(aucs), "\n")
```

## Multicollinearity Check (VIF After Ridge)

```{r}
X <- as.matrix(keep_num(X_train))
r_xx  <- cor(X)
c_val <- best_lambda

VIF <- solve(r_xx + c_val * diag(1, nrow(r_xx))) %*%
       r_xx %*%
       solve(r_xx + c_val * diag(1, nrow(r_xx)))

par(mar = c(9, 4, 4, 2))
vif_vals <- sort(diag(VIF), decreasing = TRUE)
bp4 <- barplot(vif_vals, names.arg = rep("", length(vif_vals)),
               main = "VIF After Ridge",
               ylab = "VIF",
               col = ifelse(vif_vals > 10, "tomato", "steelblue"))
text(bp4, par("usr")[3] - 0.3, labels = names(vif_vals),
     srt = 45, adj = 1, xpd = TRUE, cex = 0.55)
abline(h = 10, col = "darkred", lty = 2)
```

## Test Set Evaluation

```{r}
final_preds <- predict(fit, newx = X_te_scaled,
                       type = "response", s = best_lambda)

test_auc <- auc(roc(as.numeric(y_test[[1]]), as.numeric(final_preds), quiet = TRUE))
cat("Test AUC:", test_auc, "\n")
```

## Visualizations

**ROC curve** — shows the tradeoff between true positive rate and false positive rate on the test set.

```{r}
roc_obj <- roc(as.numeric(y_test[[1]]), as.numeric(final_preds), quiet = TRUE)
plot(roc_obj, main = paste("ROC Curve (AUC =", round(test_auc, 3), ")"))
```

**Coefficient plot** — shows which features the model relies on most at the best lambda.

```{r}
coefs <- coef(fit, s = best_lambda)
coef_df <- data.frame(
  feature = rownames(coefs)[-1],
  coef    = as.numeric(coefs)[-1]
)
coef_df <- coef_df[order(abs(coef_df$coef), decreasing = TRUE), ]
coef_df <- head(coef_df, 15)

par(mar = c(9, 4, 4, 2))
bp3 <- barplot(coef_df$coef, names.arg = rep("", nrow(coef_df)),
               main = "Top 15 Coefficients at Best Lambda",
               ylab = "Coefficient",
               col = ifelse(coef_df$coef > 0, "steelblue", "tomato"))
text(bp3, par("usr")[3] - 0.003, labels = coef_df$feature,
     srt = 45, adj = 1, xpd = TRUE, cex = 0.6)
abline(h = 0)
```

**Predicted probability distribution by fraud label** — shows how well the model separates fraud from non-fraud.

```{r}
pred_df <- data.frame(
  prob  = as.numeric(final_preds),
  label = factor(as.numeric(y_test[[1]]), labels = c("Not Fraud", "Fraud"))
)
boxplot(prob ~ label, data = pred_df,
        main = "Predicted Fraud Probability by True Label",
        ylab = "Predicted Probability", col = c("steelblue", "tomato"))
```

