---
title: "Time Series Analysis for Fraud Detection"
author: "Ovunc Ozgun Eker"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This document is the third layer of the fraud-detection pipeline. The first two layers
(provider-level ridge logistic regression and claim-level z-score ensemble) operated on
cross-sectional features. This layer asks a different question: **do fraudulent providers
show unusual temporal billing patterns** — spikes, ramps, or irregularities in their
month-by-month reimbursement totals that would not appear in clean providers?

---

## Stage 1 — Load and combine claims

```{r load}
library(tidyverse)
library(lubridate)
library(zoo)

ip     <- read_csv("Train_IP.csv",  show_col_types = FALSE) |> mutate(claim_type = 1L)
op     <- read_csv("Train_OP.csv",  show_col_types = FALSE) |> mutate(claim_type = 0L)
labels <- read_csv("Train.csv",     show_col_types = FALSE)

claims <- bind_rows(ip, op) |>
  mutate(ClaimStartDt = ymd(ClaimStartDt))

cat("Claims:", nrow(claims), "\n")
cat("Date range:", format(min(claims$ClaimStartDt, na.rm = TRUE)),
    "to", format(max(claims$ClaimStartDt, na.rm = TRUE)), "\n")
```

---

## Stage 2 — Aggregate to provider x month

```{r aggregate}
monthly <- claims |>
  mutate(month = floor_date(ClaimStartDt, "month")) |>
  group_by(Provider, month) |>
  summarise(
    monthly_claim_count         = n(),
    monthly_total_reimbursement = sum(InscClaimAmtReimbursed, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(labels |> select(Provider, PotentialFraud), by = "Provider")

cat("Provider-month rows:", nrow(monthly), "\n")
cat("Distinct providers:", n_distinct(monthly$Provider), "\n")
```

---

## Stage 3 — Exploratory plot: fraud vs non-fraud temporal patterns

We select providers with claims in at least 6 distinct months, then sample 8 fraud and
8 non-fraud providers for a faceted line chart.

```{r exploratory-plot, fig.width=12, fig.height=9}
active_providers <- monthly |>
  group_by(Provider) |>
  filter(n() >= 6) |>
  ungroup()

set.seed(42)
sample_fraud    <- active_providers |>
  filter(PotentialFraud == "Yes") |> distinct(Provider) |> slice_sample(n = 8)
sample_nonfraud <- active_providers |>
  filter(PotentialFraud == "No")  |> distinct(Provider) |> slice_sample(n = 8)

plot_data <- active_providers |>
  filter(Provider %in% c(sample_fraud$Provider, sample_nonfraud$Provider)) |>
  mutate(Label = ifelse(PotentialFraud == "Yes", "Fraud", "Non-fraud"))

ggplot(plot_data,
       aes(x = month, y = monthly_total_reimbursement / 1e3, colour = Label)) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.2) +
  facet_wrap(~ Provider, scales = "free_y", ncol = 4) +
  scale_colour_manual(values = c("Fraud" = "tomato", "Non-fraud" = "steelblue")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title  = "Monthly Total Reimbursement: 8 Fraud vs 8 Non-Fraud Providers",
       x = NULL, y = "Monthly reimbursement ($ thousands)", colour = NULL) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "top",
        strip.text  = element_text(size = 6),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 5))
```

---

## Stage 4a — Spike detector: diagnostics on the original miscalibration

The prior run flagged 72.9% of all providers — clearly too liberal. Before recalibrating,
we look at *why*: what does the distribution of roll_z values and anomalous-month counts
actually look like?

```{r spike-diagnostics, fig.width=11, fig.height=4}
# --- compute the raw z-scores (same lagged-baseline logic as before) ---
compute_z <- function(df) {
  x             <- df$monthly_total_reimbursement
  baseline_mean <- lag(rollmeanr(x, k = 3, fill = NA), 1)
  baseline_sd   <- lag(rollapply(x, width = 3, FUN = sd,
                                 align = "right", fill = NA), 1)
  roll_z <- ifelse(!is.na(baseline_sd) & baseline_sd > 0,
                   (x - baseline_mean) / baseline_sd, NA_real_)
  df |> mutate(roll_z = roll_z,
               month_anomalous_z3 = !is.na(roll_z) & roll_z > 3)
}

monthly_z <- monthly |>
  group_by(Provider) |> filter(n() >= 6) |>
  arrange(month, .by_group = TRUE) |>
  group_modify(~ compute_z(.x)) |>
  ungroup()

provider_counts <- monthly_z |>
  group_by(Provider, PotentialFraud) |>
  summarise(anomalous_months_z3 = sum(month_anomalous_z3, na.rm = TRUE),
            .groups = "drop")

par(mfrow = c(1, 3))

# z-score distribution (cap display at 20 for readability)
z_vals <- monthly_z$roll_z[!is.na(monthly_z$roll_z) &
                            monthly_z$roll_z <= 20]
hist(z_vals, breaks = 80, col = "steelblue", border = "white",
     main = "Distribution of roll_z (capped at 20)",
     xlab = "Rolling z-score", ylab = "Count")
abline(v = 3, col = "tomato", lty = 2, lwd = 2)

# anomalous-month count per provider
hist(provider_counts$anomalous_months_z3, breaks = 30,
     col = "steelblue", border = "white",
     main = "Anomalous months per provider (z > 3)",
     xlab = "Number of anomalous months", ylab = "Providers")

# how flag rate changes with min_anomalous_months threshold
thresh_seq <- 1:10
flag_rates <- sapply(thresh_seq, function(k) {
  mean(provider_counts$anomalous_months_z3 >= k)
})
plot(thresh_seq, flag_rates * 100, type = "b", pch = 16,
     col = "steelblue", xlab = "Min anomalous months required",
     ylab = "% providers flagged",
     main = "Flag rate vs. min-months threshold")
abline(h = c(5, 15), col = "tomato", lty = 2)
legend("topright", legend = "5-15% target band",
       col = "tomato", lty = 2, cex = 0.7)
```

---

## Stage 4b — Recalibrated spike detector

From the diagnostics above, we pick the minimum anomalous-month count that places the
overall flag rate inside the 5-15% target band. This choice is based solely on
"flags a sensible minority," not on what maximises the fraud ratio.

```{r spike-recalibrated}
# find the smallest threshold that brings flag rate to <= 15%
min_months_threshold <- min(which(flag_rates <= 0.15))
cat("Min anomalous months chosen:", min_months_threshold,
    " -> flag rate:", round(flag_rates[min_months_threshold] * 100, 1), "%\n\n")

provider_spike <- provider_counts |>
  mutate(is_spike_anomalous =
           anomalous_months_z3 >= min_months_threshold)

spike_rates <- provider_spike |>
  group_by(PotentialFraud) |>
  summarise(n = n(), flagged = sum(is_spike_anomalous),
            flag_rate = mean(is_spike_anomalous), .groups = "drop")
print(spike_rates)

spike_fraud    <- spike_rates |>
  filter(PotentialFraud == "Yes") |> pull(flag_rate)
spike_nonfraud <- spike_rates |>
  filter(PotentialFraud == "No")  |> pull(flag_rate)
cat("Fraud/non-fraud ratio:", round(spike_fraud / spike_nonfraud, 2), "x\n")

ct_spike <- table(PotentialFraud    = provider_spike$PotentialFraud,
                  Spike_Anomalous   = provider_spike$is_spike_anomalous)
chi_spike <- chisq.test(ct_spike)
cat("Chi-square:", round(chi_spike$statistic, 2),
    " p-value:", format(chi_spike$p.value, digits = 3), "\n")
```

---

## Stage 4c — Alternative framing: billing ramp (linear trend slope)

The exploratory plot suggested that ramps or sustained level shifts might be more
informative than single-month spikes. For each provider we fit a simple OLS line
through their monthly reimbursement series (month index 1, 2, ... as x). The slope
tells us whether billing is trending upward or downward over time.

We flag providers whose standardised slope exceeds 2 standard deviations above the
mean across all providers — i.e., unusually steep upward ramps.

```{r ramp-detector}
fit_slope <- function(df) {
  df <- df |> arrange(month) |> mutate(t = row_number())
  fit <- lm(monthly_total_reimbursement ~ t, data = df)
  tibble(slope = coef(fit)[["t"]])
}

provider_slopes <- monthly |>
  group_by(Provider, PotentialFraud) |>
  filter(n() >= 6) |>
  group_modify(~ fit_slope(.x)) |>
  ungroup()

# standardise slopes across all providers
slope_mean <- mean(provider_slopes$slope, na.rm = TRUE)
slope_sd   <- sd(provider_slopes$slope,   na.rm = TRUE)
provider_slopes <- provider_slopes |>
  mutate(slope_z          = (slope - slope_mean) / slope_sd,
         is_ramp_anomalous = slope_z > 2)

cat("Ramp-flagged providers:",
    sum(provider_slopes$is_ramp_anomalous), "/",
    nrow(provider_slopes),
    sprintf("(%.1f%%)\n\n", mean(provider_slopes$is_ramp_anomalous) * 100))

ramp_rates <- provider_slopes |>
  group_by(PotentialFraud) |>
  summarise(n = n(), flagged = sum(is_ramp_anomalous),
            flag_rate = mean(is_ramp_anomalous), .groups = "drop")
print(ramp_rates)

ramp_fraud    <- ramp_rates |>
  filter(PotentialFraud == "Yes") |> pull(flag_rate)
ramp_nonfraud <- ramp_rates |>
  filter(PotentialFraud == "No")  |> pull(flag_rate)
cat("Fraud/non-fraud ratio:", round(ramp_fraud / ramp_nonfraud, 2), "x\n")

ct_ramp <- table(PotentialFraud  = provider_slopes$PotentialFraud,
                 Ramp_Anomalous  = provider_slopes$is_ramp_anomalous)
chi_ramp <- chisq.test(ct_ramp)
cat("Chi-square:", round(chi_ramp$statistic, 2),
    " p-value:", format(chi_ramp$p.value, digits = 3), "\n")
```

---

## Stage 5 — Robustness Check 1: scale invariance

The raw-dollar ramp signal (7x ratio) could be a size artifact: large-billing providers
happen to be fraudulent AND have steeper absolute slopes simply because their numbers are
bigger. Two scale-invariant alternatives test whether the pattern survives normalisation.

### 5a — Log-scale ramp (OLS on log1p)

```{r ramp-log}
fit_slope_log <- function(df) {
  df <- df |> arrange(month) |> mutate(t = row_number())
  fit <- lm(log1p(monthly_total_reimbursement) ~ t, data = df)
  tibble(slope_log = coef(fit)[["t"]])
}

provider_slopes_log <- monthly |>
  group_by(Provider, PotentialFraud) |>
  filter(n() >= 6) |>
  group_modify(~ fit_slope_log(.x)) |>
  ungroup()

slope_log_mean <- mean(provider_slopes_log$slope_log, na.rm = TRUE)
slope_log_sd   <- sd(provider_slopes_log$slope_log,   na.rm = TRUE)
provider_slopes_log <- provider_slopes_log |>
  mutate(slope_log_z        = (slope_log - slope_log_mean) / slope_log_sd,
         is_log_ramp_anomalous = slope_log_z > 2)

cat("Log-ramp flagged:", sum(provider_slopes_log$is_log_ramp_anomalous), "/",
    nrow(provider_slopes_log),
    sprintf("(%.1f%%)\n\n", mean(provider_slopes_log$is_log_ramp_anomalous) * 100))

log_rates <- provider_slopes_log |>
  group_by(PotentialFraud) |>
  summarise(n = n(), flagged = sum(is_log_ramp_anomalous),
            flag_rate = mean(is_log_ramp_anomalous), .groups = "drop")
print(log_rates)

log_fraud    <- log_rates |> filter(PotentialFraud == "Yes") |> pull(flag_rate)
log_nonfraud <- log_rates |> filter(PotentialFraud == "No")  |> pull(flag_rate)
cat("Fraud/non-fraud ratio:", round(log_fraud / log_nonfraud, 2), "x\n")

ct_log  <- table(PotentialFraud    = provider_slopes_log$PotentialFraud,
                 Log_Ramp_Anomalous = provider_slopes_log$is_log_ramp_anomalous)
chi_log <- chisq.test(ct_log)
cat("Chi-square:", round(chi_log$statistic, 2),
    " p-value:", format(chi_log$p.value, digits = 3), "\n")
```

---

### 5b — Mean-normalised ramp

```{r ramp-meannorm}
fit_slope_norm <- function(df) {
  df <- df |> arrange(month) |> mutate(t = row_number())
  m  <- mean(df$monthly_total_reimbursement, na.rm = TRUE)
  # divide by provider mean so slope is in units of "fraction of mean per month"
  df <- df |> mutate(y_norm = monthly_total_reimbursement / pmax(m, 1))
  fit <- lm(y_norm ~ t, data = df)
  tibble(slope_norm = coef(fit)[["t"]])
}

provider_slopes_norm <- monthly |>
  group_by(Provider, PotentialFraud) |>
  filter(n() >= 6) |>
  group_modify(~ fit_slope_norm(.x)) |>
  ungroup()

slope_norm_mean <- mean(provider_slopes_norm$slope_norm, na.rm = TRUE)
slope_norm_sd   <- sd(provider_slopes_norm$slope_norm,   na.rm = TRUE)
provider_slopes_norm <- provider_slopes_norm |>
  mutate(slope_norm_z         = (slope_norm - slope_norm_mean) / slope_norm_sd,
         is_norm_ramp_anomalous = slope_norm_z > 2)

cat("Norm-ramp flagged:", sum(provider_slopes_norm$is_norm_ramp_anomalous), "/",
    nrow(provider_slopes_norm),
    sprintf("(%.1f%%)\n\n", mean(provider_slopes_norm$is_norm_ramp_anomalous) * 100))

norm_rates <- provider_slopes_norm |>
  group_by(PotentialFraud) |>
  summarise(n = n(), flagged = sum(is_norm_ramp_anomalous),
            flag_rate = mean(is_norm_ramp_anomalous), .groups = "drop")
print(norm_rates)

norm_fraud    <- norm_rates |> filter(PotentialFraud == "Yes") |> pull(flag_rate)
norm_nonfraud <- norm_rates |> filter(PotentialFraud == "No")  |> pull(flag_rate)
cat("Fraud/non-fraud ratio:", round(norm_fraud / norm_nonfraud, 2), "x\n")

ct_norm  <- table(PotentialFraud      = provider_slopes_norm$PotentialFraud,
                  Norm_Ramp_Anomalous  = provider_slopes_norm$is_norm_ramp_anomalous)
chi_norm <- chisq.test(ct_norm)
cat("Chi-square:", round(chi_norm$statistic, 2),
    " p-value:", format(chi_norm$p.value, digits = 3), "\n")
```

---

## Stage 6 — Robustness Check 2: window-edge artifact

The data spans 2008-01 through 2009-12. The first and last months may be partial (billing
not yet fully submitted or adjudicated), which could inflate the apparent ramp by creating
an artificial low start or high finish. We restrict to 2008-12 through 2009-11 — 12
complete interior months — and re-run the best ramp version.

```{r ramp-edgetrim}
monthly_trim <- monthly |>
  filter(month >= as.Date("2008-12-01"),
         month <= as.Date("2009-11-01"))

cat("Trimmed provider-month rows:", nrow(monthly_trim), "\n")
cat("Date range:", format(min(monthly_trim$month)), "to",
    format(max(monthly_trim$month)), "\n\n")

# determine best version from Check 1 (higher ratio wins)
best_ratio_check1 <- max(log_fraud / log_nonfraud,
                         norm_fraud / norm_nonfraud)

if ((log_fraud / log_nonfraud) >= (norm_fraud / norm_nonfraud)) {
  cat("Best Check-1 method: log-scale ramp\n\n")
  fit_best <- fit_slope_log
  slope_col <- "slope_log"
} else {
  cat("Best Check-1 method: mean-normalised ramp\n\n")
  fit_best <- fit_slope_norm
  slope_col <- "slope_norm"
}

provider_slopes_trim <- monthly_trim |>
  group_by(Provider, PotentialFraud) |>
  filter(n() >= 6) |>
  group_modify(~ fit_best(.x)) |>
  ungroup()

slope_trim_vals <- provider_slopes_trim[[slope_col]]
slope_trim_mean <- mean(slope_trim_vals, na.rm = TRUE)
slope_trim_sd   <- sd(slope_trim_vals,   na.rm = TRUE)
provider_slopes_trim <- provider_slopes_trim |>
  mutate(slope_trim_z        = (slope_trim_vals - slope_trim_mean) / slope_trim_sd,
         is_trim_ramp_anomalous = slope_trim_z > 2)

cat("Edge-trimmed ramp flagged:", sum(provider_slopes_trim$is_trim_ramp_anomalous), "/",
    nrow(provider_slopes_trim),
    sprintf("(%.1f%%)\n\n", mean(provider_slopes_trim$is_trim_ramp_anomalous) * 100))

trim_rates <- provider_slopes_trim |>
  group_by(PotentialFraud) |>
  summarise(n = n(), flagged = sum(is_trim_ramp_anomalous),
            flag_rate = mean(is_trim_ramp_anomalous), .groups = "drop")
print(trim_rates)

trim_fraud    <- trim_rates |> filter(PotentialFraud == "Yes") |> pull(flag_rate)
trim_nonfraud <- trim_rates |> filter(PotentialFraud == "No")  |> pull(flag_rate)
cat("Fraud/non-fraud ratio:", round(trim_fraud / trim_nonfraud, 2), "x\n")

ct_trim  <- table(PotentialFraud      = provider_slopes_trim$PotentialFraud,
                  Trim_Ramp_Anomalous  = provider_slopes_trim$is_trim_ramp_anomalous)
chi_trim <- chisq.test(ct_trim)
cat("Chi-square:", round(chi_trim$statistic, 2),
    " p-value:", format(chi_trim$p.value, digits = 3), "\n")
```

---

## Stage 7 — Full comparison table

```{r summary-table}
# original spike (z>3, any month) — recompute rate from provider_counts
orig_rates <- provider_counts |>
  group_by(PotentialFraud) |>
  summarise(flag_rate = mean(anomalous_months_z3 >= 1), .groups = "drop")
orig_fraud    <- orig_rates |> filter(PotentialFraud == "Yes") |> pull(flag_rate)
orig_nonfraud <- orig_rates |> filter(PotentialFraud == "No")  |> pull(flag_rate)
ct_orig  <- table(PotentialFraud = provider_counts$PotentialFraud,
                  Flag = provider_counts$anomalous_months_z3 >= 1)
chi_orig <- chisq.test(ct_orig)

summary_tbl <- tibble(
  Method = c(
    "Original spike (z>3, >=1 month)",
    paste0("Recalibrated spike (z>3, >=", min_months_threshold, " months)"),
    "Ramp — raw dollars (OLS slope z>2)",
    "Ramp — log1p scale",
    "Ramp — mean-normalised",
    paste0("Ramp — edge-trimmed (best of log/norm, 2008-12 to 2009-11)")
  ),
  `Overall flag %` = c(
    sprintf("%.1f%%", mean(provider_counts$anomalous_months_z3 >= 1) * 100),
    sprintf("%.1f%%", flag_rates[min_months_threshold] * 100),
    sprintf("%.1f%%", mean(provider_slopes$is_ramp_anomalous) * 100),
    sprintf("%.1f%%", mean(provider_slopes_log$is_log_ramp_anomalous) * 100),
    sprintf("%.1f%%", mean(provider_slopes_norm$is_norm_ramp_anomalous) * 100),
    sprintf("%.1f%%", mean(provider_slopes_trim$is_trim_ramp_anomalous) * 100)
  ),
  `Fraud flag %` = c(
    sprintf("%.1f%%", orig_fraud * 100),
    sprintf("%.1f%%", spike_fraud * 100),
    sprintf("%.1f%%", ramp_fraud * 100),
    sprintf("%.1f%%", log_fraud * 100),
    sprintf("%.1f%%", norm_fraud * 100),
    sprintf("%.1f%%", trim_fraud * 100)
  ),
  `Non-fraud flag %` = c(
    sprintf("%.1f%%", orig_nonfraud * 100),
    sprintf("%.1f%%", spike_nonfraud * 100),
    sprintf("%.1f%%", ramp_nonfraud * 100),
    sprintf("%.1f%%", log_nonfraud * 100),
    sprintf("%.1f%%", norm_nonfraud * 100),
    sprintf("%.1f%%", trim_nonfraud * 100)
  ),
  `Fraud/NF ratio` = c(
    round(orig_fraud / orig_nonfraud, 2),
    round(spike_fraud / spike_nonfraud, 2),
    round(ramp_fraud / ramp_nonfraud, 2),
    round(log_fraud / log_nonfraud, 2),
    round(norm_fraud / norm_nonfraud, 2),
    round(trim_fraud / trim_nonfraud, 2)
  ),
  `p-value` = c(
    format(chi_orig$p.value, digits = 3),
    format(chi_spike$p.value, digits = 3),
    format(chi_ramp$p.value, digits = 3),
    format(chi_log$p.value, digits = 3),
    format(chi_norm$p.value, digits = 3),
    format(chi_trim$p.value, digits = 3)
  )
)

knitr::kable(summary_tbl, caption = "Temporal detector comparison — all methods")
```
